Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-26 survey/survey.tex
progress on intro

No doubt, this consensus has been successful. And no doubt, these
operations \emph{can} indeed be performed on a computer. Yet, today,
with a possible quantum computing revolution in sight and with
unprecedented explosion in embedded computers and cyber-physical
systems, there are reasons to re-think this foundational question
again. In fact, the calls to re-think this foundational question have
been proclaimed by physicists for more than thirty years as the
following two quotes testify:

\begin{quote}
  Mathematical models of computation are abstract constructions, by
  their nature unfettered by physical laws. However, if these models
  are to give indications that are relevant to concrete computing,
  they must somehow capture, albeit in a selective and stylized way,
  certain general physical restrictions to which all concrete
  computing processes are subjected~\cite{toffoli:1980}.

  Another thing that has been suggested early was that natural laws
  are reversible, but that computer rules are not. But this turned out
  to be false; the computer rules can be reversible, and it has been a
  very, very useful thing to notice and to discover that. This is a
  place where the relationship of physics and computation has turned
  itself the other way and told us something about the possibilities
  of computation. So this is an interesting subject because it tells
  us something about computer rules
  \ldots~\cite{springerlink:10.1007/bf02650179}.
\end{quote}

These quotes by Toffoli and Feynman both highlight two obvious
observations: (i) all the operations that a computer performs reduce
to basic physical operations; and (ii) there is a mismtach between the
logical operations of a typical model of computation (which are
logically irreversible) and the fundamental laws of physics (which are
reversible). One could certainly dismiss the mismatch as irrelevant to
the practice of computing but our thesis is that there is the next
computing revolution is likely to be founded on revised models of
computation that are designed to in harmony with the laws of physics.

After a detailed introduction on the origins of ``logically
reversibile computer operations'' and an excursion in the origins of
our the status quo based on ``irreversible computer operations'', we
will develop in detail three reversible models of computation and
discuss their potential applications.

@@

%%%%%%%%%
\subsection{Maxwell's Demon}

To fully appreciate one of the principles is missing in conventional
computing, one must go back to an old thought experiment by
J. C. Maxwell. This is codified in a letter that Maxwell wrote to P. G
Tait in 1867 -- the letter, whose ideas are now known as
\emph{Maxwell's Demon}, tells of a thought experiment that seems to
indicate that intelligent beings can somehow violate the second law of
thermodynamics, thereby violating physics itself. Many resolutions
were offered for this conundrum (for a compilation, see Maxwell's
Demon~\cite{leff1990}), but none withstood careful scrutiny until the
establishment of \emph{Landauer's Principle} in 1961 -- a principle
whose experimental validation happened recently in
2012~\cite{berut2012experimental}.

Maxwell's demon appears to violate the second law of thermodynamics by
having a tiny `intelligence' observing the movement of individual
particles of a gas and separating fast moving particles from slow
moving ones, thereby reducing the total entropy of the
system. Landauer's resolution of the demon relied on two ideas that
took root only a few decades earlier -- the formal notion of
computation (through the work of Turing and Church, 1936) and the
formal notion of information (through the work of Claude Shannon,
1948). Landauer reasoned that the computation done by the finite brain
of the demon, involves getting information about the movement of
molecules, acting on that information and then overwriting it to make
room for the next computation.  Landauer reasoned, and this is the
important part, that the computation that is manipulating information
in the demon's brain \textit{must be thermodynamic work}, thereby
bringing the demon back into the fold of physics.

This is a strange and wonderful idea: information, physics and
computation are inextricably linked. Furthermore, this implies that
ideas in each field have consequences for the
other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,dblp:conf/csfw/malacarias12}.
%% (Cite Bennett and the various `thermodynamics of computation'
%% papers here.)

When the early models computation, the Turing machine, and the
$\lambda$-calculus, were developed, there was no compelling reason to the
take the information content of computations into
consideration -- in fact, at that time there was no quantifiable
notion of information. These models followed in the footsteps of logic
where conventional gates such as AND and OR happily erase their
inputs. These models were already radical enough, by being purely
constructive, in an era where classical mathematics, with its
pervasive use of excluded middle and frequent use of the axiom of
choice, did not overly worry about effective representation.

But first, we need to better understand our current models of
computation, as a prelude to understand where they lead us astray.

%%%%%%%%%
\subsection{Origins of Computational Models}

Aspects of computation are
quite old. M\"{u}ller (1786) first conceived of the idea of a
``difference machine'', which Babbage (1819--1822) was able to
construct. There are other computer precursors as well -- the first
stored programs were actually for looms, most notably those of
Bouchon (1725) which operated on a paper tape, and Jacquard (1804)
which operated by chains of punched cards.

@@

%%%%%%%%%
\subsection{From Classical to Quantum}

Conventional classical models of computation, including boolean
circuits, the Turing machine, and the $\lambda$-calculus, are founded on
primitives which correspond to \emph{irreversible} physical processes which
gratuitously erase information.  For example, a \emph{nand} gate is an
irreversible logical operation in the sense that its inputs cannot generally
be recovered from observing its output, and so is the operation of overriding
a cell on a Turing machine tape with a new symbol, and so is a
$\beta$-reduction which typically erases or duplicates values in a way that
is destructive and irreversible.

@@

%%%%%%%%%
\subsection{Conservation of Information}

To make this idea concrete and to provide a taste of the applications
it opens, consider a tiny 2-bit password = \verb|"10"|. The password
checker looks like:

@@

%%%%%%%%%
\subsection{Information}

We can now give a first precise definition of information preservation.
Let
`$b$' be a (not necessarily finite) type whose values are labeled
$b^1, b^2, \ldots$. Let $\xi$ be a random variable of type $b$ that is
equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
as $-\sum p_i \log{p_i}$~\cite{Shannon1948}.  Consider a function
$f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
type whose values are labeled $b_2^1, b_2^2, \ldots$. The output
entropy of the function is given by $- \sum q_j \log{q_j}$ where $q_j$
indicates the probability of the output of the function to have value
${b_2}^j$. We say a function is \emph{information-preserving} if its
output entropy is equal to the entropy of its input. In later sections,
we will refine this view.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Below are pieces of text that are not immediately useful above, but
%% from which we might still be able to pilfer usefully.
\subsection{Technical Overview}

>>>>

No doubt, this consensus has been successful. And no doubt, these
operations \emph{can} indeed be performed on a computer. Yet, today,
with a possible quantum computing revolution in sight and with
unprecedented explosion in embedded computers and cyber-physical
systems, there are reasons to re-think this foundational question
again. In fact, the calls to re-think this foundational question have
been proclaimed by physicists more than forty years ago as the
following two quotes testify:

\begin{quote}
  \textbf{Toffoli 1980~\cite{toffoli:1980}:} Mathematical models of computation are
  abstract constructions, by their nature unfettered by physical
  laws. However, if these models are to give indications that are
  relevant to concrete computing, they must somehow capture, albeit in
  a selective and stylized way, certain general physical restrictions
  to which all concrete computing processes are
  subjected.

  \textbf{Feynman 1982~\cite{springerlink:10.1007/bf02650179}:}
  Another thing that has been suggested early was that natural laws
  are reversible, but that computer rules are not. But this turned out
  to be false; the computer rules can be reversible, and it has been a
  very, very useful thing to notice and to discover that. This is a
  place where the relationship of physics and computation has turned
  itself the other way and told us something about the possibilities
  of computation. So this is an interesting subject because it tells
  us something about computer rules.
\end{quote}

\noindent These quotes by Toffoli and Feynman both highlight the
consequences of two obvious observations: (i) all the operations that
a computer performs reduce to basic physical operations; and (ii)
there is a mismatch between the logical operations of a typical model
of computation (which are logically irreversible) and the fundamental
laws of physics (which are reversible). One could certainly dismiss
the mismatch as irrelevant to the practice of computing but our thesis
is that the next computing revolution is likely to be founded on
revised models of computation that are designed to in harmony with the
laws of physics.

After a detailed introduction on the origins of ``logically
reversibile computer operations'' and an excursion into the origins of
``irreversible computer operations'', we will develop in detail three
reversible models of computation and discuss their potential
applications.

@@

%%%%%%%%%
\paragraph*{Maxwell's Demon.}
To fully appreciate the missing principle of ``reversibility'' in
conventional computing, we go back to an old thought experiment by
J. C. Maxwell. The details are codified in a letter that Maxwell wrote
to P. G Tait in 1867 -- the letter, whose ideas are now known as
\emph{Maxwell's Demon}, tells of a thought experiment that seems to
indicate that intelligent beings can somehow violate the second law of
thermodynamics, thereby violating physics itself. Many resolutions
were offered for this conundrum (for a compilation, see the book by
Leff and Rex~\cite{leff1990}), but none withstood careful scrutiny
until the establishment of \emph{Landauer's Principle} in 1961 -- a
principle whose experimental validation happened recently in
2012~\cite{berut2012experimental}.

Maxwell's demon appears to violate the second law of thermodynamics by
having a tiny `intelligence' observing the movement of individual
particles of a gas and separating fast moving particles from slow
moving ones, thereby reducing the total entropy of the
system. Landauer's resolution of the demon relied on two ideas that
took root only a few decades earlier: the formal notion of computation
(through the work of Turing and Church, 1936) and the formal notion of
information (through the work of Shannon, 1948). Landauer reasoned
that the computation done by the finite brain of the demon, involves
getting information about the movement of molecules, acting on that
information, and then overwriting it to make room for the next
computation.  In other words, the computation that is manipulating
information in the demon's brain \textit{must be thermodynamic work},
thereby bringing the demon back into the fold of physics.

%% (Cite Bennett and the various `thermodynamics of computation'  
%% papers here.)

This is a strange and wonderful idea: information, physics, and
computation are inextricably linked. In contrast, when the early
models of computation were developed, there was no compelling reason
to the take the information content of computations into consideration
-- in fact, at that time there was no quantifiable notion of
information. These models followed in the footsteps of logic where,
following hundreds of years of tradition, conventional operations such
conjunction and disjunction are logically irreversible. These models
were already radical enough, by being purely \emph{constructive}, in
an era where classical mathematics, with its pervasive use of excluded
middle and frequent use of the axiom of choice, did not overly worry
about effective representation. Landauer's observation implied however
that ideas in each field have consequences for the
other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,baez2011physics,dblp:conf/csfw/malacarias12}. To
really appreciate this fact, we delve deeper into the origin of our
computational models and argue that they are essentially reflections
of contemporary laws of physics.

\amr{wavefront}

%%%%%%%%%
\paragraph*{Origins of Computational Models.}
Aspects of computation are
quite old. M\"{u}ller (1786) first conceived of the idea of a
``difference machine'', which Babbage (1819--1822) was able to
construct. There are other computer precursors as well -- the first
stored programs were actually for looms, most notably those of
Bouchon (1725) which operated on a paper tape, and Jacquard (1804)
which operated by chains of punched cards.

@@

%%%%%%%%%
\paragraph*{From Classical to Quantum.}
Conventional classical models of computation, including boolean
circuits, the Turing machine, and the $\lambda$-calculus, are founded on
primitives which correspond to \emph{irreversible} physical processes which
gratuitously erase information.  For example, a \emph{nand} gate is an
irreversible logical operation in the sense that its inputs cannot generally
be recovered from observing its output, and so is the operation of overriding
a cell on a Turing machine tape with a new symbol, and so is a
$\beta$-reduction which typically erases or duplicates values in a way that
is destructive and irreversible.

@@

%%%%%%%%%
\paragraph*{Conservation of Information.}
To make this idea concrete and to provide a taste of the applications
it opens, consider a tiny 2-bit password = \verb|"10"|. The password
checker looks like:

@@

%%%%%%%%%
\paragraph*{Information.}
We can now give a first precise definition of information preservation.
Let
`$b$' be a (not necessarily finite) type whose values are labeled
$b^1, b^2, \ldots$. Let $\xi$ be a random variable of type $b$ that is
equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
as $-\sum p_i \log{p_i}$~\cite{Shannon1948}.  Consider a function
$f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
type whose values are labeled $b_2^1, b_2^2, \ldots$. The output
entropy of the function is given by $- \sum q_j \log{q_j}$ where $q_j$
indicates the probability of the output of the function to have value
${b_2}^j$. We say a function is \emph{information-preserving} if its
output entropy is equal to the entropy of its input. In later sections,
we will refine this view.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Below are pieces of text that are not immediately useful above, but
%% from which we might still be able to pilfer usefully.

\subsection{Technical Overview}
